Exploring Zstandard user-provided dictionary compression for FASTA files

### 2022-12-09 paper/paper.qmd
Major changes, added Zstd dictionary functions

---
title: "Exploring Zstandard user-provided dictionary compression for FASTA files"
format:
  plos-pdf:
    number-sections: false
    journal:
      # This is an identifier for the target journal: 
      # from https://plos.org/resources/writing-center/ following submission guidelines link, the identifier is the part of the URL after https://journals.plos.org/<id>/s/submission-guidelines
      id: plosbiology
    include-in-header: |
      % Remove comment for double spacing
      % \usepackage{setspace} 
      % \doublespacing
author:
  - name: Michael Persico
    equal-contributor: false
    affiliations:
      - ref: aff1
    notes: |
      "Current Address: Dept/Program/Center, Institution Name, City, State, Country"
    corresponding: false
    email: correspondingauthor@institute.edu
author-notes:
  equal-contributor: These authors contributed equally to this work.
  deceased: Deceased
  group: Membership list can be found in the Acknowledgments sections
affiliations:
  - id: aff1
    name: Department of Biology, Concordia University
    city: Montreal
    state: Quebec
    country: Canada
abstract: |
  ## Background

@@

All resources for the paper are included in the public Github repository at github.com/M-PERSIC/Persico2022.git.

The model contains two pipelines, one for Zstd custom dictionary generation and another for generating transcoded compression data using Bzip2, Xz, zLib, and Zstd compressors, with a general overview in @fig-model. Julia was chosen as there is package support for working with sequence data as well as for trivial transcoding of compressed or decompressed data via TranscodingStreams.jl wherein compression algorithm implementations are loaded as codecs[@sato]. A list of all direct dependencies can be found in the repository's Project.toml file. The Julia implementation operates by first unloading a tarball containing the set of FASTA training files into a Julia artifact, or life-cycled datastore. The files are plugged into Zstd's Dictionary API to generate a custom dictionary specific to that dataset. Random FASTA data is generated using BioSequences.jl[@biosequences] and FASTX.jl[@fastx] APIs which is then transcoded into compressed formats via the CodecBzip2.jl[@codecbzip2], CodecXz.jl[@codecxz], CodecZlib.jl[@codeczlib], and CodecZstd.jl[@codeczstd] codecs for comparative analysis. Both dictionary-based and default Zstd compression are applied to determine if the former desmontrates a significant reduction in data size.

@@

@@

I would like to express my sincere gratitude to Professor David Walsh for his advice that helped shape the research as it progressed; My friends and family that supported me throughout my studies; The Julia community for their support before and throughout the writing of this paper; The Quarto developers for producing the Quarto publishing system[@Allaire_Quarto_2022] along with the PLOS template used for this paper; Luca Di Maio and contributors to the Distrobox tool for the ease of setting up the containerized development environments[@maio]; Maximiliano Sandoval and contributors to the Citations app for managing the paper's bibliography[@sandoval], and all other persons that had indirectly assisted through their programs and research.

>>>>

---
title: "Exploring Zstandard user-provided dictionary compression for FASTA files"
format:
  plos-pdf:
    number-sections: false
    journal:
      # This is an identifier for the target journal: 
      # from https://plos.org/resources/writing-center/ following submission guidelines link, the identifier is the part of the URL after https://journals.plos.org/<id>/s/submission-guidelines
      id: plosbiology
    include-in-header: |
      % Remove comment for double spacing
      % \usepackage{setspace} 
      % \doublespacing
author:
  - name: Michael A. Persico
    equal-contributor: false
    affiliations:
      - ref: aff1
    notes: |
      "Current Address: Dept/Program/Center, Institution Name, City, State, Country"
    corresponding: false
    email: correspondingauthor@institute.edu
author-notes:
  equal-contributor: These authors contributed equally to this work.
  deceased: Deceased
  group: Membership list can be found in the Acknowledgments sections
affiliations:
  - id: aff1
    name: Department of Biology, Concordia University
    city: Montreal
    state: Quebec
    country: Canada
abstract: |
  ## Background

@@

All resources for the paper are included in the public Github repository at [github.com/M-PERSIC/Persico2022.git](https://www.github.com/M-PERSIC/Persico2022.git).

The model contains two pipelines, one for Zstd custom dictionary generation and another for generating transcoded compression data using Bzip2, Xz, zLib, and Zstd compressors, with a general overview in @fig-model. Julia was chosen as there is package support for working with sequence data as well as for trivial transcoding of compressed or decompressed data via TranscodingStreams.jl wherein compression algorithm implementations are loaded as codecs[@sato]. A list of all direct dependencies can be found in the repository's Project.toml file. The Julia implementation operates by first unloading a tarball containing the set of FASTA training files into a Julia artifact, or life-cycled datastore. The files are plugged into Zstd's training mode, via the Zstd command line interface --train option, to generate a custom dictionary specific to that dataset. Subsequent dictionary manipulation in Julia is based on an experimental CodecZstd.jl branch that includes functions that wrap the Zstd Dictionary API[@kittisopikul]. Random FASTA data is generated using BioSequences.jl[@biosequences] and FASTX.jl[@fastx] APIs which is then transcoded into compressed formats via the CodecBzip2.jl[@codecbzip2], CodecXz.jl[@codecxz], CodecZlib.jl[@codeczlib], and CodecZstd.jl[@codeczstd] codecs for comparative analysis. Both dictionary-based and default Zstd compression are applied to determine if the former desmontrates a significant reduction in data size.

@@

Zstd's multifarious design warrants further research in how tuning the implementation by altering available options (block size, compression level,...) that might influence compression performance. The development of novel data toolsets have utilized Zstd in unique ways. Aside from the Nucleotide Archival Format, FASTAFS was recently introduced as a filesystem in userspace (FUSE) toolkit wherein FASTA files are pooled into a virtualization abstraction layer that archives the set of data synced with its metadata, generating a high-level interface for accessing and storing FASTA data[@fastafs]. The zstd-seekable compression library is an embedded component that keeps the FASTAFS archive compressed for storage reduction purposes, and it represents an alternative compressed data format that deviates from Zstandard, albeit in a compatible manner, that allows for partial decompression of subranges of data[@facebook]. This is achieved via independently compressed/decompressed Zstd frames, which FASTAFS exploits for fast random access to FASTA files by decompressing only data of interest. Further areas of research for general lossless data compression in the context of biological data could include exploring other properties like compression/decompression speed performance, meaning how fast data transcoding occurs, or piloting proof-of-concept algorithms that attempt novel strategies.

@@

I would like to express my sincere gratitude to Professor David Walsh for teaching the bioinformatics course; My friends and family that supported me throughout my studies; The Julia community for their support before and throughout the writing of this paper; Mark Kittisopikul for developing the CodecZstd.jl #dictionary_and_parameters branch for Zstd dictionary functions; The Quarto developers for producing the Quarto publishing system[@Allaire_Quarto_2022] along with the PLOS template used for this paper; Luca Di Maio and contributors to the Distrobox tool for the ease of setting up the containerized development environments[@maio]; Maximiliano Sandoval and contributors to the Citations app for managing the paper's bibliography[@sandoval], and all other persons that had indirectly assisted through their programs and research.

# Conflict of interest

The author declares no conflict of interest.

### 2022-12-09 paper/paper.tex
Major changes, added Zstd dictionary functions

% Header and Footer with logo
\usepackage{lastpage,fancyhdr}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[numbers,square,comma]{natbib}
\bibliographystyle{plos2015}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Exploring Zstandard user-provided dictionary compression for FASTA files},
  pdfauthor={Michael Persico},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Exploring Zstandard user-provided dictionary compression
for FASTA
files} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
\\
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
Michael Persico\textsuperscript{1}
\\
\bigskip
\textbf{1} Department of Biology, Concordia
University, Montreal, Quebec, Canada, 
\bigskip

@@

All resources for the paper are included in the public Github repository
at github.com/M-PERSIC/Persico2022.git.

The model contains two pipelines, one for Zstd custom dictionary
generation and another for generating transcoded compression data using
Bzip2, Xz, zLib, and Zstd compressors, with a general overview in
Fig~\ref{fig-model}. Julia was chosen as there is package support for
working with sequence data as well as for trivial transcoding of
compressed or decompressed data via TranscodingStreams.jl wherein
compression algorithm implementations are loaded as codecs\citep{sato}.
A list of all direct dependencies can be found in the repository's
Project.toml file. The Julia implementation operates by first unloading
a tarball containing the set of FASTA training files into a Julia
artifact, or life-cycled datastore. The files are plugged into Zstd's
Dictionary API to generate a custom dictionary specific to that dataset.
Random FASTA data is generated using BioSequences.jl\citep{biosequences}
and FASTX.jl\citep{fastx} APIs which is then transcoded into compressed
formats via the CodecBzip2.jl\citep{codecbzip2},
CodecXz.jl\citep{codecxz}, CodecZlib.jl\citep{codeczlib}, and
CodecZstd.jl\citep{codeczstd} codecs for comparative analysis. Both
dictionary-based and default Zstd compression are applied to determine
if the former desmontrates a significant reduction in data size.

@@

@@

I would like to express my sincere gratitude to Professor David Walsh
for his advice that helped shape the research as it progressed; My
friends and family that supported me throughout my studies; The Julia
community for their support before and throughout the writing of this
paper; The Quarto developers for producing the Quarto publishing
system\citep{Allaire_Quarto_2022} along with the PLOS template used for
this paper; Luca Di Maio and contributors to the Distrobox tool for the
ease of setting up the containerized development
environments\citep{maio}; Maximiliano Sandoval and contributors to the
Citations app for managing the paper's bibliography\citep{sandoval}, and
all other persons that had indirectly assisted through their programs
and research.

>>>>

% Header and Footer with logo
\usepackage{lastpage,fancyhdr}
\usepackage{epstopdf}
%\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
%\setlength{\headheight}{27.023pt}
%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\today}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[numbers,square,comma]{natbib}
\bibliographystyle{plos2015}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Exploring Zstandard user-provided dictionary compression for FASTA files},
  pdfauthor={Michael A. Persico},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Exploring Zstandard user-provided dictionary compression
for FASTA
files} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
\\
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
Michael A. Persico\textsuperscript{1}
\\
\bigskip
\textbf{1} Department of Biology, Concordia
University, Montreal, Quebec, Canada, 
\bigskip

@@

All resources for the paper are included in the public Github repository
at
\href{https://www.github.com/M-PERSIC/Persico2022.git}{github.com/M-PERSIC/Persico2022.git}.

The model contains two pipelines, one for Zstd custom dictionary
generation and another for generating transcoded compression data using
Bzip2, Xz, zLib, and Zstd compressors, with a general overview in
Fig~\ref{fig-model}. Julia was chosen as there is package support for
working with sequence data as well as for trivial transcoding of
compressed or decompressed data via TranscodingStreams.jl wherein
compression algorithm implementations are loaded as codecs\citep{sato}.
A list of all direct dependencies can be found in the repository's
Project.toml file. The Julia implementation operates by first unloading
a tarball containing the set of FASTA training files into a Julia
artifact, or life-cycled datastore. The files are plugged into Zstd's
training mode, via the Zstd command line interface --train option, to
generate a custom dictionary specific to that dataset. Subsequent
dictionary manipulation in Julia is based on an experimental
CodecZstd.jl branch that includes functions that wrap the Zstd
Dictionary API\citep{kittisopikul}. Random FASTA data is generated using
BioSequences.jl\citep{biosequences} and FASTX.jl\citep{fastx} APIs which
is then transcoded into compressed formats via the
CodecBzip2.jl\citep{codecbzip2}, CodecXz.jl\citep{codecxz},
CodecZlib.jl\citep{codeczlib}, and CodecZstd.jl\citep{codeczstd} codecs
for comparative analysis. Both dictionary-based and default Zstd
compression are applied to determine if the former desmontrates a
significant reduction in data size.

@@

Zstd's multifarious design warrants further research in how tuning the
implementation by altering available options (block size, compression
level,\ldots) that might influence compression performance. The
development of novel data toolsets have utilized Zstd in unique ways.
Aside from the Nucleotide Archival Format, FASTAFS was recently
introduced as a filesystem in userspace (FUSE) toolkit wherein FASTA
files are pooled into a virtualization abstraction layer that archives
the set of data synced with its metadata, generating a high-level
interface for accessing and storing FASTA data\citep{fastafs}. The
zstd-seekable compression library is an embedded component that keeps
the FASTAFS archive compressed for storage reduction purposes, and it
represents an alternative compressed data format that deviates from
Zstandard, albeit in a compatible manner, that allows for partial
decompression of subranges of data\citep{facebook}. This is achieved via
independently compressed/decompressed Zstd frames, which FASTAFS
exploits for fast random access to FASTA files by decompressing only
data of interest. Further areas of research for general lossless data
compression in the context of biological data could include exploring
other properties like compression/decompression speed performance,
meaning how fast data transcoding occurs, or piloting proof-of-concept
algorithms that attempt novel strategies.

@@

I would like to express my sincere gratitude to Professor David Walsh
for teaching the bioinformatics course; My friends and family that
supported me throughout my studies; The Julia community for their
support before and throughout the writing of this paper; Mark
Kittisopikul for developing the CodecZstd.jl
\#dictionary\_and\_parameters branch for Zstd dictionary functions; The
Quarto developers for producing the Quarto publishing
system\citep{Allaire_Quarto_2022} along with the PLOS template used for
this paper; Luca Di Maio and contributors to the Distrobox tool for the
ease of setting up the containerized development
environments\citep{maio}; Maximiliano Sandoval and contributors to the
Citations app for managing the paper's bibliography\citep{sandoval}, and
all other persons that had indirectly assisted through their programs
and research.

\hypertarget{conflict-of-interest}{%
\section{Conflict of interest}\label{conflict-of-interest}}

The author declares no conflict of interest.
